---
title: "Evaluating Performance Measures"
output: html_notebook
---

The objective of this assignment is to compare two different classification algorithms using accuracy and performance metrics. I'm going to use K-NN and Naive Bayes to to predict the age of an abalone using abalone features. Abalone are shell fish that are popular to eat in many countries, especially raw in a sashimi spread. The rings attribute corresponds to an abalone's age in years (after the abalone reaches 1 to 1.5 years of age). The process of determining an abalone's age is tedious and time consuming, so using classification machine learning might be useful for predicting an abalone's age.

Data was taken from this website https://archive.ics.uci.edu/ml/datasets/Abalone

Here are the attribute descriptions:

Sex / nominal / -- / M, F, and I (infant) 
Length / continuous / mm / Longest shell measurement 
Diameter	/ continuous / mm / perpendicular to length 
Height / continuous / mm / with meat in shell 
Whole weight / continuous / grams / whole abalone 
Shucked weight / continuous	/ grams / weight of meat 
Viscera weight / continuous / grams / gut weight (after bleeding) 
Shell weight / continuous / grams / after being dried 
Rings / integer / -- / +1.5 gives the age in years

```{r}
x <- c(5, 4)
```


```{r}
 abalone <- read.csv(url("https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data"), header = FALSE, sep = ",")

colnames(abalone) <- c("sex", "length", 'diameter', 'height', 'whole_weight', 'shucked_wieght', 'viscera_wieght', 'shell_weight', 'rings' )
```


```{r}
summary(abalone)
```

```{r}
str(abalone)
```



```{r}
summary(abalone$rings)
```

As shown above, the "rings" variable has a range between 1-29. This is the variable that we want to predict, and predicting this many levels might not give us the insight we're looking for. I suspect that there's an optimal age range for harvesting abalones for consumption. While I don't know this age range, this project could be adjusted with the sought-after age range inserted. For now, we'll break the rings variable into 3 levels" "young" for abalones less than 8, "adult" for abalones between 8-11, and "old" for abalones older than 11. 

```{r}
abalone$rings <- as.numeric(abalone$rings)

abalone$rings <- cut(abalone$rings, br=c(-1,8,11,35), labels = c("young", 'adult', 'old'))


abalone$rings <- as.factor(abalone$rings)

summary(abalone$rings)
```


I'm going to create a couple of different classification models, and then compare them using accuracy and performance metrics. I'll start with a KNN classification algorithm. Because KNN requires all numeric variables for prediction, I'm going to remove the "sex" variable.


```{r}
z <- abalone
z$sex <- NULL
```

I'll now normalize the data using min max normalization

```{r}
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

z[1:7] <- as.data.frame(lapply(z[1:7], normalize))


summary(z$shucked_wieght)
```



Now each variable has a min of 0 and a max of 1. We'll now split the data into training and testing sets.

```{r}
ind <- sample(2, nrow(z), replace=TRUE, prob=c(0.7, 0.3))
KNNtrain <- z[ind==1,]
KNNtest <- z[ind==2,]
```

Now we run the model. I'm going to make k equal to the square root of 2918, the number of observations in the training set.


```{r}
library(class)
KNNpred <- knn(train = KNNtrain[1:7], test = KNNtest[1:7], cl = KNNtrain$rings, k = 54)
```



Let's see how the model does on the test data.

```{r}
library("gmodels")
CrossTable(x = KNNtest$rings, y = KNNpred, prop.chisq = FALSE)
```
```{r}
(328+451+97)/((84+2+26+91+21+159)+(328+451+97))
```

This KNN classifier predicted the abalone age with 69% accuracy - likely not accurate enough for an abalone harvester to trust. Before moving on to more specific accuracy and performance tests I'm going to try a smaller k value and see if it improves the accuracy.


```{r}
library(class)
KNNpred <- knn(train = KNNtrain[1:7], test = KNNtest[1:7], cl = KNNtrain$rings, k = 10)
```

```{r}
library("gmodels")
CrossTable(x = KNNtest$rings, y = KNNpred, prop.chisq = FALSE)
```

```{r}
(313+422+135)/((313+422+135)+(94+7+57+89+17+125))
```


This model has just about the same predictive power on the test set. This can also be shown in the confusion matrix below:


```{r}
library(caret)
confusionMatrix(KNNpred, KNNtest$rings)
```

The misclassification rate is 1 minus the accuracy, shown below.

```{r}
1-0.691
```


Let’s now create a naive bayes classifier for the same data.


```{r}
NBtrain <- KNNtrain
NBtest <- KNNtest
```

```{r}
library(e1071)
```

```{r}
model <- naiveBayes(rings ~., data = NBtrain)
model
```

```{r}
pred <- predict(model, NBtest)
print(confusionMatrix(pred,NBtest$rings))
```


The accuracy rate for the naive bayes model predicting the test set is only about 59%, which makes the misclassification rate approx. 41%.

While it's likely that neither algorithm is adequate for predicting the abalone age, the KNN model is more accurate so far.

Let's try a bootstrapping method for further model evaluation.

```{r}
library(caret)
train_control <- trainControl(method='boot', number = 100)

trModel <- train(rings~., data = z, trControl=train_control, method="nb")
```


```{r}
print(trModel)
```

```{r}
trModel2 <- train(rings~., data = z, trControl=train_control, method="knn")
```

```{r}
print(trModel2)
```


The bootstrapping method indicates that the KNN model might be slightly more accurate for classifying the data. It was also estimated that the most effective k value for the KNN model would be 9. We used 10 for our model.

Let's now do 10-fold cross validation to evaluate the models.

```{r}
control = trainControl(method="repeatedcv", number=10, repeats=3)
model5 <- train(rings~., data = trainDF, method = "knn", preProcess="scale", trControl=control)
model5
```


```{r}
control11 = trainControl(method="repeatedcv", number=10, repeats=3)
model6 <- train(rings~., data = trainDF, method = "nb", preProcess="scale", trControl=control11)
model6
```


The 10-fold cross validation method indicates that the optimal model for KNN is one with k = 9 (same as what the bootstrap method predicted).

The 10-fold cross validation method indicates that the optimal model for Naive Bayes is a model with fL = 0, usekernal = TRUE and adjust = 1.


The cross-validation method confirms that the KNN method is more effective for this data set than Naive Bayes.

Let's create the new models with the suggested parameters.
```{r}
library(class)
KNNpred <- knn(train = KNNtrain[1:7], test = KNNtest[1:7], cl = KNNtrain$rings, k = 9)
library(caret)
confusionMatrix(KNNpred, KNNtest$rings)
```

With k = 9 the model was about 69% accurate in predicting the test data set.


```{r}
library(e1071)
```

```{r}
model <- naiveBayes(rings ~., data = NBtrain, fL = 0, usekernal = TRUE, adjust = 1)
model
```

```{r}
pred <- predict(model, NBtest)
print(confusionMatrix(pred,NBtest$rings))
```

With the suggested parameters given from the 10-fold validation, the naive bayes algorithm is about 59% accurate.

The models trained by the 10-fold validation have almost equal accuracy to the models I originally created, when testing on the test data set. My concern with this project is that the parameters I originally used didn't differ much from the suggested model in 10-fold validation. 

```{r}
control14 = trainControl(method="repeatedcv", number=10, repeats=3)
model7 <- train(rings~., data = trainDF, method = "rf", preProcess="scale", trControl=control14)
model7
```

As shown above, the optimal random forest model is expected to perform at about 67%. It seems like the machine learning algorithms are having a difficult learning enough from the abalone features to accurately predict the age of the abalone. I would suspect that an abalone harvester would want a more accurate model before he or she could trust it in a commercial setting. Therefore, better data might be necessary.
